\bf\mftitle

### 2011-01-12 doc/jInfer/full/architecture/architecture.tex
3154619 Create documentation for jInfer architecture - consistent xml representation looking, some points for vektor and xml example of tree

\textit{Note: we use the term \textbf{inference} for the act of creation of schema throughout this and other jInfer documents.}\\

@@

\section{Data structures}
\subsection{Regular expressions}

For general information on regular expressions, please refer to \cite{wikiregexp}, \cite{automatatheory}.
All classes pertaining to regular expressions can be found in the package \code{cz.cuni.mff.ksi.jinfer.base.regexp}.
In jInfer, we use extended regular expressions as they give us nicer syntax (and easier programming).

@@

Regexp instance is by default created as immutable, that is, once instantiated, you cannot add more children to list of children, cannot change type, content etc. It is to prevent missuse. In special circumstances, one does not know future children of regexp in time of creation. This occurs mainly in input modules, where by parsing XML data sequentially, one does not know contents of element in time of handling start element event.
For these cases, special \code{getMutable()} method is implemented to obtain regexp with none of members set. One has to fill in all properties carefully and call \code{setImmutable()} aftewards. Proper usage should be one of following:
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.LAMBDA);
	r.setImmutable();
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.TOKEN);
	r.setContent(...)
	r.setImmutable();
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.CONCATENATION);
	r.addChild(...);
	r.addChild(...);
	r.addChild(...);
	r.setImmutable();
\end{verbatim}

@@

@@

\subsection{(Nondeterministic) Finite Automaton}

@@

jInfer and its documentation uses extended context-free grammars. % TODO vektor Cite, for example http://www.engr.mun.ca/~theo/Courses/fm/pub/context-free.pdf
\textit{Rules} in such grammar are in the form
% TODO anti Please make a proper arrow
\[  Left Hand Side (LHS)  -> Right Hand Side (RHS) \]
where LHS is a letter of the alphabet (token), RHS is a regular expression over this alphabet. Example would be
\begin{verbatim}
  a -> b, (c | d)*
\end{verbatim}

In jInfer each such rule is represented with an \code{Element} instance. In this representation, the \code{Element} itself is the LHS, its \code{subnodes} are the RHS.\\

Another important notion is a \textit{grammar}. A grammar consists of its rules, so in jInfer a grammar is just a collection of \code{Element}s. Closely related term is \textit{Initial Grammar}, which for us is a grammar consisting of rules with \textit{simple} right hand sides, i.e. just concatenations of tokens. Initial Grammar is produced by Initial Grammar Generator. % TODO vektor Link

% TODO explain metadata and those that we already use
% TODO section about NBMs

@@

% TODO anti: is it really necessary to write fig. by hand? Shouldn't it be automatic?

The process by which jInfer infers the resulting schema from various inputs (inference process) is summarized by fig. \ref{inference_process}. From the high-level viewpoint, it consists of three consecutive steps carried out by three different modules:
\begin{enumerate}
	\item Initial Grammar (IG) generation: done by the \textit{Initial Grammar Generator (IGG)} module, this is the process of converting all of the inputs to IG representation. All documents, schemas and queries selected as input are evaluated, simple rules are extracted and in the end sent to the next step.\\
	% TODO anti: how about the whole examples in italic?
	For example, a trivial XML document
	\begin{verbatim}
  <a>
    <b>
      <c/>
    </b>
    <d/>
  </a>
	\end{verbatim}
	would translate into the following IG rules
	\begin{verbatim}
		a -> b, d
		b -> c
	\end{verbatim}	
	\item Simplification: done by the \textit{Simplifier} module, this is the process of simplifying, compressing or somehow compactly describing the IG by a smaller number of (more complex) rules. User interaction might be used in this step to help achieve better simplification. At the end of this step, all rules are sent to the last step.\\
	For example, rules
	\begin{verbatim}
		a -> b
		a -> c, d, d, d
	\end{verbatim}
	could be simplified to a single rule
	\begin{verbatim}
		a -> b | (c, d*)
	\end{verbatim}
	\item Schema export: done by the \textit{Schema Generator (SchemaGen)} module, this is the process of actually creating the resulting schema from the simplified rules. Result of this step is a string representation of the schema, which is sent back to the framework (and later displayed, saved, etc).\\
	% TODO vektor Example
\end{enumerate}
Important thing to note here is that all these steps are executed consecutively. That means, \textit{Simplifier} is only started \textit{after} the \textit{IGG} completely finished its work and returned IG to be simplified. Similarly, \textit{SchemaGen} gets all the rules to export at once, in one list.\\

This modular architecture means that (at least theoretically) it is possible to replace any and all of these modules with \textit{something} else that does similar job. In practice, it might be useful to use the logic already implemented in \textit{BasicIGG} and just extend it to handle new input type.

@@

\textit{Initial Grammar Generator} takes an object of type \textit{Input}. This class encapsulates all the input files in 3 collections of \code{File}: \code{documents}, \code{schemas} and \code{queries}. Enumerating these files provides IGG with access to all data it needs to create Initial Grammar.\\

\textit{Simplifier} and \textit{Schema Generator} take grammar, % TODO vektor Link to grammar definition
in other words a list of rules as input. In the first case, this grammar is the Initial Grammar, in the second case it is the simplified grammar.

@@

Each callback interface naturally belongs to the similarly named module interface. As their respective interfaces, also callbacks define one crucial method: \textit{finished}. Each inference module is responsible for invoking this method on the callback it got as a parameter after it has finished its work and has results to be passed on. Again, these 3 finished methods have different arguments based on the inference module.\\

@@

User running the inference might change his mind and try to stop this. For this reason, modules have to check for this case in every time-consuming place such as long loops with the following code
% TODO anti Please indent
\begin{verbatim}
  for (forever) {
    if (Thread.interrupted()) {
      throw new InterruptedException();
    }
    doStuff();
  }
\end{verbatim}

>>>>

\emph{Note: we use the term \textbf{inference} for the act of creation of schema throughout this and other jInfer documents.}\\

@@

\section{Data structures}
\subsection{Regular expressions}
% TODO anti lambda, epsilon notation
For general information on regular expressions, please refer to \cite{wikiregexp}, \cite{automatatheory}.
All classes pertaining to regular expressions can be found in the package \code{cz.cuni.mff.ksi.jinfer.base.regexp}.
In jInfer, we use extended regular expressions as they give us nicer syntax (and easier programming).

@@

Regexp instance is by default created as immutable, that is, once instantiated, you cannot add more children to list of children, cannot change type, content etc. It is to prevent missuse. In special circumstances, one does not know future children of regexp in time of creation. This occurs mainly in input modules, where by parsing XML data sequentially, one does not know contents of element in time of handling start element event.
For these cases, special \code{getMutable()} method is implemented to obtain regexp with none of members set. One has to fill in all properties carefully and call \code{setImmutable()} aftewards. Proper usage should be one of following:
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.LAMBDA);
	r.setImmutable();cz.cuni.mff.ksi.jinfer.base
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.TOKEN);
	r.setContent(...)
	r.setImmutable();
\end{verbatim}
\begin{verbatim}
	Regexp<T> r = Regexp.<T>getMutable();
	r.setInterval(...);
	r.setType(RegexpType.CONCATENATION);
	r.addChild(...);
	r.addChild(...);
	r.addChild(...);
	r.setImmutable();
\end{verbatim}

@@

\begin{figure}
\caption{XML document representation} \label{xml_example}
\centering\includegraphics[scale=\myscale]{xml_example}
\end{figure}
Let's take an example, the following XML document would be represented as on fig. \ref{xml_example}.
\begin{verbatim}
  <person name="john" surname="smith">
    <info>
      Some text
      <note/>
    </info>
    <more/>
  </person>
\end{verbatim}

@@

@@

jInfer and its documentation uses extended context-free grammars\cite{extendedcfg}. % TODO vektor seee literature.bib and repair citation http://www.engr.mun.ca/~theo/Courses/fm/pub/context-free.pdf
\emph{Rules} in such grammar are in the form
% TODO vektor read "the not so short introduction to latex" and math mode section
$$
  \textnormal{Left Hand Side (LHS)} \to \textnormal{Right Hand Side (RHS)}
$$
where LHS is a letter of the alphabet (token), RHS is a regular expression over this alphabet. Example would be
$$
  a \to b, (c | d)\ast
$$
In jInfer each such rule is represented with an \code{Element} instance. In this representation, the \code{Element} itself is the LHS, its \code{subnodes} are the RHS.\\

Another important notion is a \emph{grammar}. A grammar consists of its rules, so in jInfer a grammar is just a collection of \code{Element}s. Closely related term is \emph{Initial Grammar}, which for us is a grammar consisting of rules with \emph{simple} right hand sides, i.e. just concatenations of tokens. Initial Grammar is produced by Initial Grammar Generator. % TODO vektor Link
% TODO vektor don't forget about advanced interface by capabilities.... don't know if to be mentioned right here, but don't forget
% TODO explain metadata and those that we already use
% TODO section about NBMs (anti: but i suppose it on the end of document)

\subsection{Nondeterministic Finite Automaton}
For all inference algorithms based on merging states of NFA's, our implementation of nondeterministic finite automaton might be interesting.
Implementation consists of 4 classes: \code{Automaton<T>}, \code{Step<T>}, \code{State<T>} and \code{AutomatonCloner<A, B>} (see package \code{cz.cuni.mff.ksi.jinfer.base.automaton}).
Whole implementation uses java generics for representation of symbol of alphabet, we denote by \code{T} the java type of symbol.

Let's begin by smalles of the classes, the \code{State<T>}.
It represents automaton state, it has its \code{name} member (integer)

@@

% TODO anti: is it really necessary to write fig. by hand? Shouldn't it be automatic? anti: don't know, who cares... once it figure, once picture and maybe painting :D

The process by which jInfer infers the resulting schema from various inputs (inference process) is summarized by fig. \ref{inference_process}. From the high-level viewpoint, it consists of three consecutive steps carried out by three different modules:
\begin{enumerate}
	\item Initial Grammar (IG) generation: done by the \emph{Initial Grammar Generator (IGG)} module, this is the process of converting all of the inputs to IG representation. All documents, schemas and queries selected as input are evaluated, simple rules are extracted and in the end sent to the next step.\\
	% TODO anti: how about the whole examples in italic? anti: bleee
	For example, a trivial XML document
	\begin{verbatim}
  <a>
    <b>
      <c/>
    </b>
    <d/>
  </a>
	\end{verbatim}
	would translate into the following IG rules
	\begin{eqnarray*}
		a & \to &  b, d \\
		b & \to & c \\
		c & \to & \lambda \\
		d & \to & \lambda 
	\end{eqnarray*}
	\item Simplification: done by the \emph{Simplifier} module, this is the process of simplifying, compressing or somehow compactly describing the IG by a smaller number of (more complex) rules. User interaction might be used in this step to help achieve better simplification. At the end of this step, all rules are sent to the last step.\\
	For example, rules
	\begin{eqnarray*}
		a & \to & b \\
		a & \to & c, d, d, d
	\end{eqnarray*}
	could be simplified to a single rule
	\begin{eqnarray*}
		a & \to & b | (c, d\ast)
	\end{eqnarray*}
	\item Schema export: done by the \emph{Schema Generator (SchemaGen)} module, this is the process of actually creating the resulting schema from the simplified rules. Result of this step is a string representation of the schema, which is sent back to the framework (and later displayed, saved, etc).\\
	% TODO vektor Example
\end{enumerate}
Important thing to note here is that all these steps are executed consecutively. That means, \emph{Simplifier} is only started \emph{after} the \emph{IGG} completely finished its work and returned IG to be simplified. Similarly, \emph{SchemaGen} gets all the rules to export at once, in one list.\\

This modular architecture means that (at least theoretically) it is possible to replace any and all of these modules with \emph{something} else that does similar job. In practice, it might be useful to use the logic already implemented in \emph{BasicIGG} and just extend it to handle new input type.

@@

\emph{Initial Grammar Generator} takes an object of type \emph{Input}. This class encapsulates all the input files in 3 collections of \code{File}: \code{documents}, \code{schemas} and \code{queries}. Enumerating these files provides IGG with access to all data it needs to create Initial Grammar.\\

\emph{Simplifier} and \emph{Schema Generator} take grammar, % TODO vektor Link to grammar definition
in other words a list of rules as input. In the first case, this grammar is the Initial Grammar, in the second case it is the simplified grammar.

@@

Each callback interface naturally belongs to the similarly named module interface. As their respective interfaces, also callbacks define one crucial method: \emph{finished}. Each inference module is responsible for invoking this method on the callback it got as a parameter after it has finished its work and has results to be passed on. Again, these 3 finished methods have different arguments based on the inference module.\\

@@

User running the inference might change his mind and try to stop this. For this reason, modules have to check for this case in every time-consuming place such as long loops with the following code
\begin{verbatim}
  for (forever) {
    if (Thread.interrupted()) {
      throw new InterruptedException();
    }
    doStuff();
  }
\end{verbatim}
